Open-source landscape for 3D CSEM modelling

### 2020-10-19 manuscript/Manuscript.tex
Final draft for review to co-authors.

The 3D CSEM code \emg3d is a multigrid solver \citep{CMMP.64.Fedorenko} for electromagnetic diffusion following \cite{GP.06.Mulder}, including tri-axial electrical anisotropy. The matrix-free solver can be used as main solver or as preconditioner for one of the Krylov subspace methods implemented in SciPy. The governing equations are discretized on a staggered grid by the finite-integration technique \citep{AEU.77.Weiland}, which is a finite-volume generalisation of a Yee grid. The code is written completely in Python using the NumPy and SciPy stacks, where the most time- and memory-consuming parts are sped up through jitted (just-in-time compiled) functions using Numba \citep{LLVM.15.Lam}. The strength of \emg3d is the matrix-free multigrid implementation, which is characterised by almost linear scaling both in terms of runtime and memory usage, and it is therefore a solver that uses comparably very little memory.

@@

The electromagnetic implementations use a staggered grid finite volume approach where physical properties are discretized at cell centres, fields on cell edges and fluxes on cell faces. There are two different discretization strategies implemented for Maxwell's equations: (I) the EB-formulation, which discretizes the electric field ($\vec{e}$) and the magnetic flux density ($\vec{b}$) and (II) the HJ-formulation, which discretizes the magnetic field ($\vec{h}$) and the current density ($\vec{j}$). Having multiple implementations allows for testing that compares results from each approach, as well as the representation of both electrical and magnetic sources on cylindrically symmetric meshes. \simpeg supports variable magnetic permeability and full-tensor anisotropy for the physical properties. \simpeg interfaces to various solvers including \texttt{PARDISO} and \texttt{MUMPS}, and also to some implemented in SciPy.

@@

It is easy to run a simulation and obtain reasonable-looking results. Though we may be able to confirm the accuracy of codes in some simple scenarios, validating the performance of 3D CSEM codes for real 3D problems is only possible by cross-validating multiple solutions. Overall, we observed an excellent match between all solutions. The relative misfits, almost always smaller than 10\,\%, were related to the either weak or strong amplitudes of real or imaginary parts and can be attributed to particularities regarding the discretization, boundary effects, and interpolation strategy. Considering this result in terms of accuracy and robustness, it is more interesting to discuss the characteristics of the four considered codes with focus on the demand of computational resources and the capability of handling the investigated CSEM geometries, including the very important underlying mesh design. The following arguments are based on a comparison of only marine CSEM models, which are the focus of industry applications. Nevertheless, we are confident that many observations would hold in a similar manner for land-based, airborne, or mixed setups.

For the MR3D results one might argue that 5-10\,\% NRMSD is a lot. We argue that this is probably as good as it gets, given that the code with which we compare is not open-source. This has been shown within the process of these results. Our first attempt were based on the originally published, fine MR3D model. This model is very detailed, but its extent is too small in x- and y-direction for CSEM computation, and each of our codes did its own extrapolation. However, none of the codes were able to reproduce the published results. The reason is that there are many ways how you can extrapolate a model. We reached out to the authors of MR3D, and they kindly agreed to also published the upscaled and extended computational model. It is only with this model that we were able to get comparable results.

We point out that making a synthetic and a real benchmark model available to cross-validate any 3D CSEM code was the determining reason for their choice, regardless of the suitability of the considered FV and FE codes for these problems. The first model, including the block-anomalies variation, is exactly reproducible by using any numerical method with either regular or irregular computational domains. Aside from the idea of analysing an industrially relevant benchmark example, the MR3D model was also considered since the primary design from \cite{GEO.19.Correa} uses a regular discretization. At least theoretically, the exact geometry might be reconstructed with unstructured meshes, but it is never possible to convert an irregular geometry to a regular grid without approximations. Hence, the chosen examples favoured the FV codes, since they used either regular grids or octree-meshes, whereas the two FE codes were required to tetrahedralize the model geometries.

In the first model, the necessity of discretizing regular structures with tetrahedra, especially thin layers, led to comparatively large FE problem sizes and computational resource demands. FE codes with unstructured meshes were probably the poorest choice of all commonly used methods for this setup in terms of computational performance. Similar challenges arise when considering an octree mesh. However, if we introduce just one slight irregularity such as bathymetry, a thin dipping plate or an anticline, the performance behaviour could likely switch. All of these changes would barely change the problem size for the FE codes, but significantly increase it for the FV codes if the true geometry should be approximated sufficiently accurate.

The real MR3D model was, ironically, the ideal case for FE codes because it's structure was characterised by irregular lithological horizons obtained from seismic data. Since the corresponding published resistivity model was defined on a regular grid, we were forced to approximate the comparatively fine regular discretization by an unstructured one and interpolate the resistivity data for being able to apply the FE codes to this problem. We are completely aware about the ineptness of this re-approximation procedure, unless it served for the cross-validation purposes. Nevertheless, the chosen tetrahedral discretization, especially in combination with second order basis functions, showed clearly the intrinsic strength of the FE codes, requiring significantly less dof for meshing irregular geometries in general. The p2 FE solution required only $\approx$1.9\,M dof instead of $\approx$90\,M dof for the FV system by \citet{GEO.19.Correa}, disregarding the structure of the system matrices. It is important to state that the FE mesh used for the MR3D model incorporates the entire domain and could be used for all source and receiver positions, whereas the rectilinear and octree meshes were designed for the shown receiver and two source lines.

The codes under consideration use direct and iterative solvers. In general, the biggest advantage of iterative solvers with appropriate preconditioners are the comparatively small memory requirements. Direct solvers can be considered as most robust to solve any ill-conditioned SLE, which is most important for systems of the FE method on unstructured meshes. As the consumed computational resources indicate, a more powerful machine than a laptop is required for most 3D CSEM problems, but no high-performance-computing architecture.

The computation times can differ significantly for specific simulations, but there is no clear general advantage for one of the solver types. In our models, the runtime comparison favoured the iterative solver and therefore, \emg3d. Even though not shown here, direct solvers exhibit their strongest advantage, besides stability, in terms of computation times if responses of multiple CSEM transmitters need to be computed in the same computational domain. As the system matrix factorisation requires 98-99\,\% of the solution time, computations for additional sources come at almost no cost for direct solvers, i.e., MUMPS, whereas it would come at the same cost as the first source for \emg3d. For the sake of completeness, note that there are also advanced techniques for iterative solvers to make use of preconditioners, factorisations, or intermediate solutions of the previous SLE to speed up the computation for multiple sources, but this argumentation is beyond the scope of this work.

We see this work as the start for future elaboration on the validation of modelling codes, not necessarily restricted to CSEM problems. Much more comparisons and examples are required which not only result in the finding that a sufficient accuracy was observed. 3D CSEM modelling is a difficult task, which requires many considerations. It starts with the selection of the right code for the problem. We found that the meshing task, which is most performance determining, is particularly difficult and relevant, choosing cells small enough to appropriately represent the model yet to be as coarse as possible still achieving the desired precision. The required model extent has to be considered, as do the effects of the so-called airwave in marine setups or boundary meshes in general. This is not a new finding but rather well known fact. We are surprised that only the minority of publications in the field of 3D EM modelling considers a sufficiently detailed elaboration of the meshing process.

A completely objective evaluation of codes is only possible by cross-validation. Nevertheless, we point out that using different discretizations or formulations of the numerical approach, e.g. by choosing field or potential approaches, second or total field formulations, different polynomial order basis functions, and others, is a suitable alternative to the cross-comparison of multiple codes for self-validating complex 3D results. This is a very powerful method to confirm the general functionality of codes for any modelling problem, which might not be covered by existing cross-validation benchmark examples. Considering for instance  this study, it would be very optimistic to estimate the accuracy of the utilised codes for a land-based mineral exploration setup with a highly-conductive, steeply-dipping conductor. However, each modeller could run multiple simulations with different configurations to increase the reliability of the obtained results.

It is worth noting that the reported runtime is just one of the aspects, and we would like to emphasise that neither the runtime nor the memory consumption are at the core of this comparison, but the validation of the results. As such, no special efforts were undertaken to minimise either, as this is an entire different task. Another aspect is to generate an appropriate mesh; with all the testing involved, this step can easily take days of work. As such the computation times becomes relative, and the ease of usability of a code has to be considered too. For \custem, \petgem, and \simpeg it is therefore also a question of how easy everything is to set-up, because after the set-up phase it depends largely on the solver used, which might change in the future; \simpeg, e.g., is looking at also interfacing to similar solvers as currently are used by \custem and \petgem. It is, however, slightly different for \emg3d, as that is primarily a solver on its own. It is also important to state that while \emg3d, \custem, and \petgem are purely CSEM codes, \simpeg is a larger framework which includes other geophysical methods as well.

The examples shown consider the geophysical problem of marine CSEM with resistive bodies. We hope to see in the future similar comparisons for other cases, such as land CSEM with strong topography looking for conductive bodies. The codes we demonstrated could be, and many frequently are, applied to very different applications, including any other geophysical exploration in the sea or on land (oil, mining, geothermal, groundwater, environmental) and also non-geophysical tasks. The current main applications of the different codes are: mining and environmental (\custem, \simpeg), oil and more recent geothermal (\petgem), and oil, methane, and deep sea mining (\emg3d).

A positive outcome of a collaboration between different projects such as this is that it brings the realm as a whole further, which should be motivation enough for further collaborations. Within \simpeg, this work has motivated feature development including averaging strategies to map physical properties on a fine mesh to a coarser mesh for computation and new examples to be included in the documentation for designing octree meshes. Furthermore, as a part of a broader development objective of inter-operating with other forward-simulation engines, connecting \emg3d and \simpeg provided a motivating use-case for the latest refactor and release of \simpeg 0.14.0. Within \emg3d, this work pushed a lot of the meshing functionality, and implementing of I/O utilities for different file formats. Within \custem, cell-wise resistivity interpolation was added for this work, multi-layer subsurface topography was the first time applied as well as first time reciprocity modelling, and also improved mesh design (resolution, refinement, etc.). Within \petgem, this work has promoted the inclusion of magnetic field computation, improving the generation of adapted meshes based on a semi-automatic approach that reduces user intervention. Also added was the continuous integration functionality and improved its test suite.

\section{Conclusions \& Outlook}

We compared four different open-source 3D CSEM modellers by computing responses for a layered, anisotropic model, a model containing three resistive blocks, and the realistic marine Marlim R3D model. All comparisons exhibited an excellent reliability of the solutions, in the order of $\approx1\,\%$ relative error for the layered and the block model, and $\approx10\,\%$ NRMSD for the real world model. Our data should make it very easy for new codes to have a readily available data set to test against and validate. We are confident that the discussions about runtimes, memory consumption, solver choices, and, most important, discretizations of the 3D CSEM problems help potential users not only to decide which modeller is best suited for specific tasks, but also to support their individual projects.

As none of the considered methods is best suited for all problems, it is important to have more test models for various scenarios in future. Our study is limited to marine CSEM cases in the frequency-domain. Reasonable extensions include providing benchmark models for land, airborne, or mixed CSEM environments as well as time-domain data. Another addition to this work would be focusing on complex irregular  models, tailored for an FE or FV code based on unstructured meshes, and to compare how the FV codes can cope with it.

We encourage the community to not only work on new open-source code developments but also to create a landscape of easily accessible benchmark models for increasing the number of reliable and reproducible solutions. Our study could provide one of the primary inputs for this task. We are confident that the mentioned development will and should be taking place, taking into account the current general trend in science to open-source papers, codes or data.

Our study is one example of a collaboration facilitated by open practices including sharing of code and data. We hope that these results may be useful for the entire CSEM community at large, and we invite and encourage the community to make more code, modelling scripts, and results publicly available.

@@

We had help from several colleagues for different tasks, and would like to thank Bane Sullivan (plotting of tetrahedra meshes in matplotlib through PyVista), Seogi Kang (input and octree mesh design for the \simpeg layer and block models), and Joseph Capriotti (octree mesh implementation and volume averaging in \simpeg). We would also like to thank Paulo Menezes for the help and explanations with regards to the Marlim R3D model and corresponding CSEM data, and for making their actual computation model available under an open-source license.

>>>>

The 3D CSEM code \emg3d is a multigrid solver \citep{CMMP.64.Fedorenko} for electromagnetic diffusion following \cite{GP.06.Mulder}, including tri-axial electrical anisotropy. The matrix-free solver can be used as main solver or as preconditioner for one of the Krylov subspace methods implemented in SciPy. The governing equations are discretized on a staggered grid by the finite-integration technique \citep{AEU.77.Weiland}, which is a finite-volume generalisation of a Yee grid. The code is written completely in Python using the \texttt{NumPy} and \texttt{SciPy} stacks, where the most time- and memory-consuming parts are sped up through jitted (just-in-time compiled) functions using \texttt{Numba} \citep{LLVM.15.Lam}. The strength of \emg3d is the matrix-free multigrid implementation, which is characterised by almost linear scaling both in terms of runtime and memory usage, and it is therefore a solver that uses comparably very little memory.

@@

The electromagnetic implementations use a staggered grid finite volume approach where physical properties are discretized at cell centres, fields on cell edges and fluxes on cell faces. There are two different discretization strategies implemented for Maxwell's equations: (I) the EB-formulation, which discretizes the electric field ($\vec{e}$) and the magnetic flux density ($\vec{b}$) and (II) the HJ-formulation, which discretizes the magnetic field ($\vec{h}$) and the current density ($\vec{j}$). Having multiple implementations allows for testing that compares results from each approach, as well as the representation of both electrical and magnetic sources on cylindrically symmetric meshes. \simpeg supports variable magnetic permeability and full-tensor anisotropy for the physical properties. \simpeg interfaces to various solvers including \texttt{PARDISO} and \texttt{MUMPS}, and also to some implemented in \texttt{SciPy}.

@@

%% 0. What we did and why %%
Today we are in the fortunate situation where we have several open-source codes available to model CSEM data for arbitrarily sized complex 3D models. It is therefore comparably easy for anyone to run a simulation and obtain reasonable-looking, robust and precise results, something that was impossible a few years ago. However, just because one obtains a good-looking result does not say anything about its accuracy. Though we may be able to confirm the accuracy of codes in some simple scenarios, validating the performance of 3D CSEM codes for complex problems is only possible by cross-validating multiple solutions. This is the core motivation for this study.

%% 1. Results: good agreement %%
Overall, we observed an excellent match between the solutions. The amplitudes of the layered model have mostly a relative error of less than $1-2\,$\%. Looking individually at the real or imaginary part results in much higher error values, but those are related to the zero crossings of the responses. The cross-validation of the results for the block model yields a similar picture, with NRMSD values within a few percents. For the large, complex MR3D model it is a bit different. First it can be noted that the errors seem to be higher for higher frequencies. For the lowest frequency, 0.125\,Hz, the NRMSD is a few percent for inline and broadside $E_x$ components, and up to 10\,\% for the broadside $E_y$ component. For higher frequencies these numbers rise to the order of $\pm10\,$\%. The reason for the higher misfit for MR3D than for the layered or block model lies mostly in the discretization, boundary effects, and interpolation strategy. While one might argue that $5-10\,$\% NRMSD is a lot, we argue that this is probably as good as it gets, given that the code with which the responses we compare to were computed is not open-source. This has been shown within the process of this study. Our first attempt were based on the originally published, fine MR3D model. This model is very detailed, but its extent is too small in x- and y-directions for CSEM computations, and each of our codes did its own extrapolation. However, none of the codes were able to reproduce the published results. The reason is that there are many ways how you can extrapolate a model. We reached out to the authors of MR3D, and they kindly agreed to also published the upscaled and extended computational model. It is only with this model that we were able to get comparable results.

%% 2. Discretization %%
One of the key insights we gained from this study is that creating appropriate meshes is a difficult and time-consuming task. While there exist now several open-source 3D CSEM codes the capability for automatic meshing lacks behind and currently remains a largely manual task requiring experience in the field. It is important to state that the chosen models favour structured meshes. The first two examples can easily and accurately be represented by regular meshes. Using, for instance, a dipping layer instead of three resistive cubes would turn it around and make the model more favourable for unstructured meshes. MR3D would be ideal an scenario for unstructured meshes, as it is based on also published irregular lithological horizons. Since the corresponding resistivity model, however, was defined on a regular grid, we were forced to approximate the comparatively fine regular discretization by an unstructured one and interpolate the resistivity data for being able to apply the FE codes to this problem. We are completely aware about the ineptness of this re-approximation procedure, unless it served for the cross-validation purposes. Both cases show that automatic and adaptive meshing capabilities are needed, which will increase the flexibility of 3D CSEM modelling. Such meshing codes should also take into account the physics of CSEM computation with its diffusive behaviour. It is also important to state that the unstructured mesh used for the MR3D model incorporates the entire domain and could be used for all source and receiver positions, whereas the rectilinear and octree meshes were designed for the shown receiver and corresponding source lines.

%% 3. Iterative vs direct solvers; runtime and memory %%
Whereas proper discretization is the main driver for accurate results, the used solver is the main discriminator in terms of runtime and memory. Both iterative and direct solvers are used in this study, however, the chosen models clearly favour iterative solvers. The biggest advantage of iterative solvers are the comparatively small memory requirements, even more so if the iterative solver is matrix free. Direct solvers exhibit their strongest advantage, besides stability and robustness, in terms of computation times if responses of multiple CSEM transmitters need to be computed in the same computational domain. As the system matrix factorisation requires 98-99\,\% of the solution time, computations for additional sources come at almost no cost for direct solvers, whereas it would come at the same cost as the first source for iterative solvers. Independently of the solver there is a trade-off between high accuracy and runtime. As an example, the octree mesh made for a very fast computation in the MR3D example thanks to the low number of dof, even though a direct solver was used. But the octree mesh has, due to its 2:1 aspect ratio, some limitations on the accuracy, why it was not used in the first two examples. Having discussed the discretization and the solver it is important to state that the reported runtime and memory is just one of the aspects, and we would like to emphasise that neither was at the core of this comparison, but the validation of the results. As such, no special efforts were undertaken to minimise either, as this is an entire different task.

%% 4. Focus here on marine CSEM; gains through collaboration %%
The shown examples consider the geophysical problem of marine CSEM with resistive bodies in the frequency domain. We hope to see in the future similar comparisons for other cases, such as land CSEM with strong topography; airborne and mixed air-ground surveys; looking for conductive bodies; time domain; and models based on horizons instead of blocks which pose a challenge for structured grids and favour unstructured meshes. A positive outcome of a collaboration between different projects such as this is that it brings the realm as a whole further, which should be motivation enough for further collaborations. Within \simpeg, this work has motivated feature development including averaging strategies to map physical properties on a fine mesh to a coarser mesh for computation and new examples to be included in the documentation for designing octree meshes. Furthermore, as a part of a broader development objective of inter-operating with other forward-simulation engines, connecting \emg3d and \simpeg provided a motivating use-case for the latest refactor and release of \simpeg 0.14.0. Within \emg3d, this work pushed a lot of the meshing functionality, and implementation of I/O utilities for different file formats. Within \custem, cell-wise resistivity interpolation was added for this work, multi-layer subsurface topography was the first time applied as well as first time reciprocity modelling, and also improved mesh design (resolution, refinement, etc.). Within \petgem, this work has promoted the inclusion of magnetic field computation, improving the generation of adapted meshes based on a semi-automatic approach that reduces user intervention. Also added was the continuous integration functionality and improved its test suite.

\section{Conclusions}

We compare the computed fields and underlying meshes of four different open-source 3D CSEM codes by means of three increasingly complex resistivity models. A layered marine, anisotropic model is used as first example to verify our results with semi-analytical responses. All codes show excellent accuracy with a relative error in the order of a few percents. Three resistive blocks are added to the layered model to increase complexity in the second example. The cross-validation of the outcomes of the different codes shows a NRMSD of a few percent, hence in a similar range as the relative error in the layered model. The third example is Marlim R3D, a realistic, complex, marine resistivity model. The model and corresponding CSEM responses are open-source, allowing an independent validation of our results with results from another (closed-source) code. In this scenario the NRMSD is in the order of $5-10$\,\%, where the main source can be attributed to differences in discretization. The required runtime and memory consumption is primarily controlled by the used solver. The accuracy, on the other hand, depends to a large degree on the discretization. A proper discretization is therefore key for an accurate result, which makes this step the most time-consuming task, not the actual computation of the responses itself. Validating the correctness of a 3D code is a difficult task, and it is essential to have easily accessible benchmark models with reliable and reproducible solutions. Our study is one example of a collaboration facilitated by open practices including sharing of code and data. We hope that these results may be useful for the entire CSEM community at large, and we invite and encourage the community to make more code, modelling scripts, and results publicly available.

@@

We had help from several colleagues for different tasks, and would like to thank Bane Sullivan (plotting of tetrahedra meshes in \texttt{matplotlib} through \texttt{PyVista}), Seogi Kang (input and octree mesh design for the \simpeg layer and block models), and Joseph Capriotti (octree mesh implementation and volume averaging in \simpeg). We would also like to thank Paulo Menezes for the help and explanations with regards to the Marlim R3D model and corresponding CSEM data, and for making their actual computation model available under an open-source license.
